Modal Abstractions for Virtualizing Memory Addresses

### 2024-06-30 examples.tex
+

As we explain in Section \ref{}, a traversal over a page-table tree can be done through the indices into the tables at each level followed by obtaining the address of the next table. In Figure \ref{fig:getnexttable}, we see how to obtain the physicall address of the next level page-table. Traversal start from the root address of the L4 table (\textsf{pte\_t *pml4} given as a parameter to \textsf{walkpgdir}). Then, we obtain the address of the entry in the L4 table via the index in the virtual address (\textsf{va}).

Finally, we call \textsf{get\_next\_table} to get the virtual address to the next level page-table -- L3.
Within \textsf{get\_next\_table} first checks if the entry whose address is passed as a parameter -- e.g., for L4 \textsf{pml4[PML4EX(va)]}.
In case of absence, \textsf{pte\_initialize} is called for a physical-zeroed page (\textsf{kalloc} -- currently an axiom as it is not part of \textsf{pte} module) to be backing the physical frame address of the entry.
\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
\[
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner\urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$

@@

Once we make sure that the frame of the page-table entry is backed with an allocated physical page, we can obtain the virtual address of the next level-page table (\textsf{next} of function \textsf{get\_next\_table} in Figure \ref{}) via looking up the \textit{identity mapping} from the physical address (the address value referenced by \textsf{entry->pfn}) to virtual address value to be returned by \textsf{get\_next\_table}.
\subsubsection{Identity Mappings}
Most kernels maintain an invariant on their page tables that the virtual address of any page used for a page table lives at a virtual address
whose value is a constant offset from the physical address a practice sometimes referred to as \emph{identity mapping} (even though the physical-to-virtual translation
is typically not literally the identity function, but adding a constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.}
\newcommand{\qfraczero}{\textsf{qfrac}}
\newcommand{\true}{\textsf{true}}
\tikzstyle{boxedassert_border} = [sharp corners,line width=0.2pt]
\NewDocumentCommand \boxedassertpv {O{} m o}{%
	\tikz[baseline=(m.base)]{
		%	  \node[rectangle, draw,inner sep=0.8pt,anchor=base,#1] (m) {${#2}\mathstrut$};
		\node[rectangle,inner sep=1.5pt,outer sep=0.2pt,anchor=base] (m) {${\,#2\,}\mathstrut$};
		\draw[#1,boxedassert_border] ($(m.south west)$) rectangle ($(m.north east)$);
	}\IfNoValueF{#3}{^{\,#3}}%
}
\newcommand*{\knowInvpv}[2]{\boxedassertpv{#2}[#1]}
\newcommand*{\ownGhostpv}[2]{\boxedassertpv[dash dot]{#2}[#1]}

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
  \end{figure*}

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
\[
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner\urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$

@@

>>>>

\subsubsection{Loading Page-Table Address Value}
In Figure \ref{fig:getnexttable}, we see how to obtain the physical address of the next level page-table. Traversal start from the root address of the page-table tree (the address of L4 table \textsf{pte\_t *pml4}) given as a parameter to \textsf{pte\_walkpgdir}). Then, we obtain the address of the entry (\textsf{pml4[PML4EX(va)]}) in the L4 table via the index in the virtual address (\textsf{va}).

Finally, we call \textsf{get\_next\_table} to get the virtual address to the next level page-table -- at this point the address of L3 table.
Within \textsf{get\_next\_table} first checks if the entry whose address is passed as a parameter ($\textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry}$). To be able to do so, it needs to load the entry value (a 64-bits word divided into bit-fields for corresponding fileds shown in Section {}) and check that \textit{presence} bit of the entry is set or not.
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape]
;;if(entry->present)
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{\_} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$
mov    -0x8[rbp],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$ 
mov     rdi, r8
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$ 
mov    [r8],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$ 
  \end{lstlisting}
\vspace{-1em}
\caption{Making sure that \textsf{entry}'s frame is referencing to an address backed up by a physical page.}
\label{fig:loadentry}
\end{figure}
As we see in Figure \ref{fig:loadentry}, we have the physical address (\textsf{entry}) of the page table entry kept in the register $r8$, but not the memory mapping to load the entry value referenced by the physical address \textsf{entry}. 
\subsubsection{Identity Mappings}
Most kernels maintain an invariant on their page tables that the virtual address of any page used for a page table lives at a virtual address
whose value is a constant offset from the physical address a practice sometimes referred to as \emph{identity mapping} (even though the physical-to-virtual translation
is typically not literally the identity function, but adding a constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.}
\newcommand{\qfraczero}{\textsf{qfrac}}
\newcommand{\true}{\textsf{true}}
\tikzstyle{boxedassert_border} = [sharp corners,line width=0.2pt]
\NewDocumentCommand \boxedassertpv {O{} m o}{%
	\tikz[baseline=(m.base)]{
		%	  \node[rectangle, draw,inner sep=0.8pt,anchor=base,#1] (m) {${#2}\mathstrut$};
		\node[rectangle,inner sep=1.5pt,outer sep=0.2pt,anchor=base] (m) {${\,#2\,}\mathstrut$};
		\draw[#1,boxedassert_border] ($(m.south west)$) rectangle ($(m.north east)$);
	}\IfNoValueF{#3}{^{\,#3}}%
}
\newcommand*{\knowInvpv}[2]{\boxedassertpv{#2}[#1]}
\newcommand*{\ownGhostpv}[2]{\boxedassertpv[dash dot]{#2}[#1]}

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
  \end{figure*}
Now, to read the value referenced by physical address \textsf{entry} in Figure \ref{fig:loadentry}, first, we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and obtain the virtual mapping out of the identity mapping $\Xi$ with the side condition
\[
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner
 \]
 assuring that looking up identity mappings are safe, because the full ownership asserts the non-presence of the entry which requires allocation a new physical page and setting the presence bit. By obtaining the identity mapping for \textsf{entry} in Figure \ref{fig:loadentry} -- ($\textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val}$ with page addr \textsf{entry}),
 \[
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{entry\},m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$ 
mov    [r8],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast  \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$ 
 \]
 we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit via $and 0x1,rdi$.

 \subsubsection{Ensuring Presence of Entry}
In case of absence, \textsf{pte\_initialize} is called for a physical-zeroed page (\textsf{kalloc} -- currently an axiom as it is not part of \textsf{pte} module but the allocator's) to be backing the physical frame address of the entry.

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner\urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$

@@

Once we make sure that the frame of the page-table entry is backed with an allocated physical page, we can obtain the virtual address of the next level-page table (\textsf{next} of function \textsf{get\_next\_table} in Figure \ref{}) via looking up the \textit{identity mapping} from the physical address (the address value referenced by \textsf{entry->pfn}) to virtual address value to be returned by \textsf{get\_next\_table}.
\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
\[
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner\urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$

@@

\subsubsection{}
